Capon Beamforming Applied on Moving Objects - An Analysis of Local Lateral Shift-Invariance

### 2013-12-05 HRUIofMovingObjects/jrnl_article.tex
Moving objects: Finally all sections are completed. Let the feedback loop begin!

\begin{abstract}
%\boldmath
If an ultrasound imaging system provides a presentation of a moving object which is sensitive to local spatial shifts, the system is said to be spatially locally shift-variant. This can for instance happen if the axial or lateral sampling is insufficient. The Capon beamformer has been shown to provide increased lateral resolution in ultrasound images. Increased lateral resolution should demand denser lateral sampling. In previous literature on Capon beamforming for medical ultrasound imaging however only single frame scenarios have been simulated, and temporal behaviour and effects caused by the increased resolution and lack of oversampling have therefore been neglected.

@@

\IEEEPARstart{I}{n} the last decade several authors have applied the Capon beamformer (also known as the minimum variance beamformer) for ultrasound imaging \cite{Synnevag2007, Vignon2008, Viola}. The adaptive nature of the method has been shown to greatly improve the lateral resolution in ultrasound images. In recent work we introduced a fast implementation of the Capon beamformer, capable of processing a stream of images at interactive frame rates \cite{Asen}. It is therefore an interesting question how a moving object appears after Capon beamforming has been applied. Earlier work on Capon beamforming for medical ultrasound imaging has investigated the method's capabilities in single frames only, and temporal behaviour and effects has therefore been neglected.

An important property of an ultrasound imaging system is its local shift-invariance. Local shift-invariance means that an object's appearance should be preserved when the object, or the ultrasound probe, is moved slightly. This invariance for local movements is what makes ultrasound imaging feasible in general, but it is specially important for quantitative techniques that correlate time-varying features, like pulsed doppler and flow imaging, and speckle tracking of tissue and blood. In this paper we analyze the lateral shift-invariance property of the Capon beamformer by studying the beamformer output when applied on simulated and \textit{in vitro} images of moving objects.

As mentioned, Capon beamforming has been shown to provide improved lateral resolution and edge definitions in ultrasound images \cite{Synnevag2007, Synnevag2009, Chen2011}. The method accomplishes this by adjusting the array weights based on element-to-element covariance. Hence, the method yields data dependent weights. Capon beamforming is however also known for its suboptimal performance if the sample size available for covariance estimation is limited \cite{Mestre2006}, if correlated signals are present \cite{Widrow1982}, and if the assumed steering vector does not match the signal propagation vector \cite{Wax1996, Wax1996a}. This has led to the development of several robust versions of the Capon beamformer with e.g. spatial smoothing \cite{Shan1985} or diagonal loading \cite{JianLi2003} of the covariance estimate, or by applying steering vector uncertainty sets (SVUS) \cite{Lorenz2005, Rubsamen2013}. The first two, together with temporal smoothing of the covariance estimate, have been applied in earlier work on Capon beamforming for medical ultrasound imaging \cite{Synnevag2009}. The first two trade resolution for robustness and temporal smoothing helps to preserve speckle statistics for high resolution settings \cite{Synnevag2007a}. SVUS reduces the signal-nulling effect of the Capon beamformer which appears if the signal is present during the covariance estimation process and if this estimate is later used with a steering vector that does not match the signal propagation vector exactly. This self-nulling is actually how the method obtains its super resolution capabilities, but the signal can also get lost in the process. In this paper we make use of a steering vector set to reduce this self-nulling effect.

In \cite{Cox1973}, Cox derives formulas for the resolution of the Capon beamformer, and presents in great detail how the amount of mismatch between the selected steering vector and the signal propagation vector impacts the beamformer output in simple situations. Cox states the following important fact about the Capon beamformer (the $\vec{k}_3$ processor): 
\begin{quote}
"\textit{... a $\vec{k}_3$ processor requires more closely spaced beams than a ... $\vec{k}_1$-processor in order to avoid serious signal suppression effects being introduced on signals arriving from directions between the beams}", 
\end{quote}
where $\vec{k}_1$ is the delay-and-sum (DAS) beamformer. In Fig.\,\ref{fig:das_capon_beams} we have plotted the two-way response for DAS and the Capon beamformer using (26) and (32) in \cite{Cox1973}. A 64-element linear array with $\lambda/2$ element spacing, where $\lambda$ is the wavelength, is scanned over a sector equal to two times the two-way system resolution ($1/64$ radians) using a narrow band pulse. Located in the farfield of the sector is an object moving parallel to the array surface. The plot shows the scanned output when the object is located at plus and minus half the system resolution. No spatially white noise is present. Hence, the interference-noise matrix ($\sigma_0^2\mat{Q}$ in \cite{Cox1973}) is the identity. The effect of typical subarray averaging and diagonal loading ($L=32$ and $d=1/100$) has also been added to the Capon beamformer output. A diagonal loading of $1/100$ corresponds to an signal-to-noise ratio of 20 dB before array gain is applied. The figure shows that a DAS beamformer, with a beamspacing equal to the system resolution, could yield a 1.9 dB scalloping loss \cite{Harris1978} if the object is moving parallel to the array. However if the Capon beamformer is used with the same beam density the scalloping loss could be as large as $27$ dB.

\begin{figure}[!t]
\centerline{
\includegraphics[width=\linewidth]{\img fig1_2way.eps}
}
\caption{Theoretical drop in output power mid-way between receive beams for delay-and-sum and Capon beamforming ($M=64$, $L=32$, $d=0.01$) applied in a farfield-narrowband setting. The scanned output from both beamformers is plotted for two different time instances. For the first instances an object is located in the farfield of the scanned sector at an angle equal to half the system resolution, and for the latter the object is moved a distance corresponding to the system resolution.}
\label{fig:das_capon_beams}
\end{figure}

@@

In this paper we investigate the issue further, and we show that when a moving point scatterer is imaged with the Capon beamformer with the same beam spacing sufficient for DAS the predicted scalloping loss is observed, manifested as blinking point scatteres. The question is then, what can we do in order to improve on the situation without affecting the imaging frame rate and processing throughout. In this paper we therefore study the scalloping loss of the Capon beamforming and investigate different methods aimed at reducing it. We also discuss how the different methods will affect the imaging frame rate and the processing complexity.

In the next section we give an introduction to Capon beamforming for medical ultrasound imaging together with a short introduction to lateral sampling and spatial shift invariance. In Section \ref{sec:capon_LLSI} the local lateral shift-invariance of the Capon beamformer is investigated, before the different methods are presented in Section \ref{sec:methods}. Then, in Section \ref{sec:res} and Section \ref{sec:dis} we present and discuss the results of applying the most promising method on \textit{in-vitro} data. Finally we give our conclusions in Section \ref{sec:con}.

@@

The lateral sampling required in an ultrasound image is given by the combined aperture size on transmit ($D_{tx}$) and receive ($D_{rx}$), and the center frequency ($f_c = c/\lambda$), where $c$ is the speed of sound in the body and $\lambda_c$ is the wavelength of the center frequency. The required Nyquist beam spacing is then \cite{Hergum2007}:
\begin{align}
\Delta < \frac{\lambda_c}{D_{tx} + D_{rx}}. \label{eq:resolution}
\end{align}
If this requirement is not fulfilled the scalloping loss could become visible in the image and the system will no longer be locally laterally shift-invariant. In an ultrasound image with 256 gray levels and 50 dB dynamic range, a 1 dB loss corresponds to 5 gray levels. This is approximately equal to the visibility threshold (Weber fraction of 2\%) for a grayscale image, and a loss larger than 1 dB could therefore end up being visible to the observer. In our analysis a system would, based on this reasoning, be regarded as locally laterally shift-invariant if the loss in between beams is smaller than 1 dB. However, depending on viewing light conditions and e.g. the selected gray level mapping, a larger loss might be acceptable.

In order to visualize the local lateral shift-invariance of a given imaging system, we will make use of lateral shift-variance plots (LSV-plots), introduced by Hergum \textit{et al.} \cite{Hergum2007}. A LSV-plot is constructed for a given imaging system by imaging a point scatterer moving laterally at constant speed. For each lateral position the root-mean-square (RMS) of the image data is calculated in range, and the resulting beam profiles are stacked in to an image. For a given scan sequence there are twice as many point positions than there are receive lines. The LSV-plots in this paper are displayed as a contour plot with contours on -1, -2, -3, -6, -12 and -24 dB. The maximum for each beam profile is also marked with a black dot, and the amplitude variation among these points is presented in a subplot next to the LSV-plot. \comment{Add interp factor}

To further quantify the shift-invariance we have calculated the mean absolute error (MAE) of the maximum point position in each beam profile versus a straight line ($f(x)=x$). We will refer to this measure as P-MAE. The MAE of the maximum point's amplitude versus the data maximum is also measured. We refer to this measure as A-MAE. The first measure quantifies the amount of geometric distortion \cite{Hergum2007}, and the latter quantifies the peak gain variation or in our case the scalloping loss. The maximum amplitude deviation, or the maximum scalloping loss, along this point trace is also calculated. We present all these numbers together with the LSV-plots.

A given imaging system will then be locally laterally shift-invariant if the contours are diagonal and if the amplitude deviation among the maximum points is less than 1 dB. Then there is minimal geometric distortion (low P-MAE) and the scalloping loss will hopefully not be visible for the observer (low A-MAE). %Since both effects can occur at the same time it is important to have a measure not effected by geometric distortion.

Figure \ref{fig:das_shiftinvariant} presents such a LSV-plot for an imaging system using the DAS beamformer with a beam spacing equal to (\ref{eq:resolution}). The red dotted vertical lines indicates the position of transmit-receive beam pairs. The point scatterer response has been simulated using Field II \cite{Jensen1992, Jensen1996a} with a probe setup equal to the M5S probe by GE Vingmed Ultrasound, Horten, Norway. The probe was excited with a 2.5 MHz and 1.5 cycles long pulse with a $\text{f-number}=3$ and corresponding cosine apodization on receive. The transmit focus in azimuth and elevation was configured to be equal and the point scatterer was moved through this focus. The lateral sampling density has been doubled prior to detection through coherent interpolation. We see that the contours in Fig.\,\ref{fig:das_shiftinvariant} are approximately diagonal (low P-MAE) and that the amplitude variation among the maximum points is minimal (low A-MAE). We therefore conclude that the system is locally laterally shift-invariant. In this, and all the following LSV-plots based on simulations, the scan sequence always has a beam at zero degrees and the point scatterer is located exactly on this beam at one time instance.

The scalloping loss in Fig.\,\ref{fig:das_shiftinvariant} is smaller than what was predicted by Fig.\,\ref{fig:das_capon_beams}. To understand why note that Fig.\,\ref{fig:das_capon_beams} shows the scanned response for a narrowband system while the imaging system in Fig.\,\ref{fig:das_shiftinvariant} uses a broadband pulse. The beam in Fig.\,\ref{fig:das_shiftinvariant} while therefore be more similar to a squared gaussian function than the squared $sinc$ functions depicted in Fig.\,\ref{fig:das_capon_beams}. The imaging system in Fig.\,\ref{fig:das_shiftinvariant} also uses an expanding cosine window on receive.

@@

\begin{figure*}[!t]
\centerline{
\subfloat[]{\includegraphics[width=0.5\linewidth]{\img fig2a.eps}\label{fig:das_shiftinvariant}}
\hfill{}
\subfloat[]{\includegraphics[width=0.5\linewidth]{\img fig2b.eps}\label{fig:das_shiftvariant}}
}
\caption{LSV-plots of delay-and-sum beamforming. If the contours are diagonal and the amplitude variation among the peak points (black dots) is small, the imaging system is said to be locally laterally shift-invariant. Red vertical lines are plotted where transmit-receive beam pairs are located. a) Sampling corresponding to (\ref{eq:resolution}). The system is locally laterally shift-invariant. b) Undersampling with a factor of two. The system is no longer locally laterally shift-invariant.}
\label{fig:das}
\end{figure*}

\section{Local Lateral Shift-Invariance of the Capon Beamformer}\label{sec:capon_LLSI}
We will now investigate the local lateral shift-invariance of the Capon beamformer with parameters equal to the ones used in earlier work \cite{Synnevag2009, Asen}. In Fig.\,\ref{fig:capon1} the same point scatterer response as presented in Fig.\,\ref{fig:das_shiftinvariant} has been weighted with Capon weights calculated using (\ref{eq:capon_weights}) with $L = M/2 = 32$, $K=1$ and $d=1/100$. As mentioned, the set of lateral displacements is constructed in such a way that the center beam is the only one hitting the point scatterer exactly. The signal cancellation caused by the difference between the signal propagation vector and the assumed steering vector is therefore minimal at this location. From the plot we observe that the signal cancellation is high in all the other lateral positions. As a consequence the scalloping loss is tremendous, and not far from the predicted value in Fig\,\ref{fig:das_capon_beams}. %The difference between the predicted and the simulated loss might be caused by several factors. The most important one being that Fig.\,\ref{fig:das_capon_beams} shows a single frequency whereas the RMS operator used in Fig.\,\ref{fig:capon1} average a broadband signal in range. For a slice through the center of the pulse the scalloping loss is below -30dB.

@@

\begin{figure*}[!t]
\centerline{
\subfloat[]{
\includegraphics[width=0.5\linewidth]{\img fig3a.eps}\label{fig:capon1}
}
\subfloat[]{
\includegraphics[width=0.5\linewidth]{\img fig3b.eps}\label{fig:capon2}
}
}
\caption{LSV-plots of Capon beamforming with $L=M/2=32$ and $K=1$, and with sampling corresponding to (\ref{eq:resolution}). a) Capon beamforming with $d=1/100$ as diagonal loading. The system is not locally lateral shift-invariant. b) The diagonal loading factor $d$ is increased to 1. The system is still not locally lateral shift-invariant.}
\label{fig:capon}
\end{figure*}

@@

\subsection{Oversampling on Transmit}
The straight forward way of obtaining sufficient sampling in an active system is by increasing the number of transmit beams (tx-beams) until the system is locally laterally shift-invariant. In Fig.\,\ref{fig:capon_oversampling_a} and Fig.\,\ref{fig:capon_oversampling_b} we have increased the lateral sampling on transmit by a factor of four and ten \comment{Increase the sampling to 16x to get the scalloping loss below 1 dB?} respectively. As expected the geometric distortion and scalloping loss decreases with increased lateral sampling. Note that Fig.\,\ref{fig:capon_oversampling_b} is zoomed compared to Fig.\,\ref{fig:capon} in order to better see the beam-to-beam variation.

@@

\begin{figure*}[!t]
\centerline{
\subfloat[]{
\includegraphics[width=0.5\linewidth]{\img fig4a.eps}\label{fig:capon_oversampling_a}
}
\subfloat[]{
\includegraphics[width=0.5\linewidth]{\img fig4b.eps}\label{fig:capon_oversampling_b}
}
}
\caption{Capon beamforming as described in Fig\,\ref{fig:capon} with oversampling on transmit. The figure is zoomed compared with Fig\,\ref{fig:capon} to better see the beam-to-beam variation. a) Four times oversampling. b) Ten times oversampling.}
\label{fig:capon_oversampling}
\end{figure*}

@@

For an uniform linear array located along the x-axis the steering vector $\vec{a}_\theta$ should be calculated as 
\begin{align}
\vec{a}_\theta = 
\begin{bmatrix}
e^{-j\frac{2\pi}{\lambda_c}x_0\sin(\theta)} \\
e^{-j\frac{2\pi}{\lambda_c}x_1\sin(\theta)} \\
\vdots \\ 
e^{-j\frac{2\pi}{\lambda_c}x_{L-1}\sin(\theta)}
\end{bmatrix}
\end{align}
where $x_i$ is the element position and $\theta$ is swiped from $-\Delta/2$ to $\Delta/2$ where $\Delta$ is as defined in (\ref{eq:resolution}). Note that the steering vector is calculated for a $L$ element array centered around the origin, even though the full array has $M$ elements. It is important that this subarray has the same element pitch, $x_i - x_{i-1}$, as the full array.

@@

Since the phase rotation method operates on the same data as before, there will be no reduction in the acquisition frame rate by applying this method. In addition, since the same number of covariance matrices are constructed and inverted, there is no extra cost involved here as well. The only additional cost will be the multiplication of the extra steering vectors with the inverted covariance matrix. If the equation $\mat{\hat{R}}\vec{a} = \vec{b}$, as in \cite{Asen}, is solved using Gauss-Jordan elimination, bringing matrix $[\mat{\hat{R}}  \vec{a}_{\theta_1} \vec{a}_{\theta_2} \cdots \vec{a}_{\theta_{N_\theta}}]$ into a reduced row echelon form has the same complexity as for inverting $\mat{\hat{R}}$, as long as $N_\theta < L$.

@@

\section{In-vitro phantom data}\label{sec:res}
In Fig.\,\ref{fig:LSV_invitro} we present the result of applying the Capon beamformer on \textit{in-vitro} data with and without oversampling by phase rotation. The data was acquired using a modified Vivid E9 high-end cardiac ultrasound scanner by GE Vingmed Ultrasound, Horten, Norway, equipped with a M5S transducer operating in harmonic mode. The transmit beam density was set to match (\ref{eq:resolution}). A tissue mimicking phantom containing wire targets and cysts was imaged, and the data was acquired while the probe was moved across the phantom manually. Fig.\,\ref{fig:image_das} shows the first frame out of 14 in the acquired loop. %Movies of the whole loop processed with DAS and Capon beamforming with and without ten times oversampling by phase rotation can be found in the supplementary material \multimedia{Videos}.

In Fig\,\ref{fig:image_das} one point scatterer has been selected (red box \comment{convert into a thicker white box}), and then used to generate LSV-plots.  \comment{Todo: Add lateral coherent interpolation as done for the simulations? As for now speckle is not fully resolved without it.}

Figure \ref{fig:phantom_das} shows the resulting LSV-plot when using the DAS beamformer. The scalloping loss is less than 1 dB, and the system is therefore locally laterally shift-invariant. As a consequence point scatterers has a consistent appearance when subjected to motion \multimedia{Video DAS}.

>>>>

\begin{abstract}
%\boldmath
If an ultrasound imaging system provides a presentation of a moving object which is sensitive to local spatial shifts, the system is said to be spatially locally shift-variant. This can for instance happen if the axial or lateral sampling is insufficient. The Capon beamformer has been shown to provide increased lateral resolution in ultrasound images. Increased lateral resolution should demand denser lateral sampling. In previous literature on Capon beamforming for medical ultrasound imaging however only single frame scenarios have been simulated. Temporal behaviour and effects caused by the increased resolution and lack of oversampling have therefore been neglected.

@@

\IEEEPARstart{I}{n} the last decade several authors have applied the Capon beamformer (also known as the minimum variance beamformer) for ultrasound imaging \cite{Synnevag2007, Vignon2008, Viola}. The adaptive nature of the method has been shown to greatly improve the lateral resolution in ultrasound images, but with a high computational cost. In recent work we introduced an implementation of the Capon beamformer capable of processing a stream of images at interactive frame rates \cite{Asen}. It is therefore an interesting question how a moving object appears after Capon beamforming has been applied. Earlier work on Capon beamforming for medical ultrasound imaging has investigated the method's capabilities in single frames only, and temporal behaviour and effects has therefore been neglected.

An important property of an ultrasound imaging system is it's local spatial shift-invariance. Local spatial shift-invariance means that an object's appearance should be preserved when the object, or the ultrasound probe, is moved slightly. This invariance to local movements is what makes ultrasound imaging feasible in general, but it is specially important for quantitative techniques that correlate time-varying features, like pulsed doppler and flow imaging, and speckle tracking of tissue and blood. In this paper we analyze the lateral shift-invariance property of the Capon beamformer by studying the beamformer output when applied on simulated and \textit{in vitro} images of moving objects.

As mentioned, Capon beamforming has been shown to provide improved lateral resolution and edge definitions in ultrasound images \cite{Synnevag2007, Synnevag2009, Chen2011}. The method accomplishes this by adjusting the array weights based on element-to-element covariance. Hence, the method yields data dependent weights. Capon beamforming is however also known for its suboptimal performance if the sample size available for covariance estimation is limited \cite{Mestre2006}, if correlated signals are present \cite{Widrow1982}, and if the assumed steering vector does not match the signal propagation vector \cite{Wax1996, Wax1996a}. This has led to the development of several robust versions of the Capon beamformer with e.g. spatial smoothing \cite{Shan1985} or diagonal loading \cite{JianLi2003} of the covariance estimate, or by applying steering vector uncertainty sets (SVUS) \cite{Lorenz2005, Rubsamen2013}. The first two, together with temporal smoothing of the covariance estimate, have been applied in earlier work on Capon beamforming for medical ultrasound imaging \cite{Synnevag2009}. The first two trade resolution for robustness and temporal smoothing helps to preserve speckle statistics for high resolution settings \cite{Synnevag2007a}. SVUS helps to reduce the signal-nulling effect of the Capon beamformer which appears if the signal is present during the covariance estimation process and if this estimate is later used with a steering vector that does not match the signal propagation vector exactly. This self-nulling is actually how the method obtains it's super resolution capabilities, but the signal can also get lost in the process. In this paper we will make use of steering vector set to reduce this self-nulling effect.

In \cite{Cox1973}, Cox derives formulas for the resolution of the Capon beamformer, and presents in great detail how the amount of mismatch between the selected steering vector and the signal propagation vector impacts the beamformer output in simple situations. Cox states the following important fact about the Capon beamformer (the $\vec{k}_3$ processor): 
\begin{quote}
"\textit{... a $\vec{k}_3$ processor requires more closely spaced beams than a ... $\vec{k}_1$-processor in order to avoid serious signal suppression effects being introduced on signals arriving from directions between the beams}", 
\end{quote}
where $\vec{k}_1$ is the delay-and-sum (DAS) beamformer. In Fig.\,\ref{fig:das_capon_beams} we have plotted the two-way response for DAS and the Capon beamformer using (26) and (32) in \cite{Cox1973}. A 64-element linear array with $\lambda/2$ element spacing, where $\lambda$ is the wavelength, is scanned over a sector equal to two times the two-way system resolution ($1/64$ radians) using narrow band excitation. Located in the farfield of the sector is an object moving parallel to the array surface. The plot shows the scanned output when the object is located at plus and minus half the system resolution. No spatially white noise is present. Hence, the interference-noise matrix ($\sigma_0^2\mat{Q}$ in \cite{Cox1973}) is the identity. The effect of typical subarray averaging and diagonal loading ($L=32$ and $d=1/100$) has also been added to the Capon beamformer output. A diagonal loading of $1/100$ corresponds to an signal-to-noise ratio of 20 dB before array gain is applied. The figure shows that a DAS beamformer, with a beamspacing equal to the system resolution, could yield a $-1.9$ dB scalloping loss \cite{Harris1978} if the object is moving parallel to the array. However if the Capon beamformer is used with the same beam density the scalloping loss could be as large as $-27$ dB. A $-1.9$ dB scalloping loss for the Capon beamforming is obtained for this specific example when the beam density is increased with a factor of 25.

\begin{figure}[!t]
\centerline{
\includegraphics[width=\linewidth]{\img fig1_2way.eps}
}
\caption{Theoretical drop in output power mid-way between receive beams for delay-and-sum and Capon beamforming ($M=64$, $L=32$, $d=0.01$) applied in a farfield-narrowband setting. The scanned output from both beamformers is plotted for two different time instances. For the first instances an object is located at an angle equal to half the system resolution, and for the latter the object is moved an angular distance corresponding to the system resolution.}
\label{fig:das_capon_beams}
\end{figure}

@@

In this paper we investigate the issue further, and we show that when a moving point scatterer is imaged with the Capon beamformer with the same beam spacing sufficient for DAS the predicted scalloping loss is observed. The question is then, what can we do in order to improve on the situation without affecting the imaging frame rate and processing throughout. In this paper we therefore study the scalloping loss of the Capon beamforming and investigate different methods aimed at reducing it. We also discuss how the different methods will affect the imaging frame rate and the processing complexity.

In the next section we give an introduction to Capon beamforming for medical ultrasound imaging together with a short introduction to lateral sampling and spatial shift invariance. In Section \ref{sec:capon_LLSI} the local lateral shift-invariance of the Capon beamformer is investigated, before different methods for improved local lateral shift-invariance are presented in Section \ref{sec:methods}. Then, in Section \ref{sec:res} we present the results of applying the most promising method on \textit{in-vitro} data. Finally we discuss our findings in and Section \ref{sec:dis} and give our conclusions in Section \ref{sec:con}.

@@

The lateral sampling required in an ultrasound image for a given center frequency, $f_c$, is given by the combined aperture size on transmit ($D_{tx}$) and receive ($D_{rx}$), and the wavelength of the center frequency, $\lambda_c = c/f_c$, where $c$ is the speed of sound in the body. The required Nyquist beam spacing is then \cite{Hergum2007}:
\begin{align}
\Delta < \frac{\lambda_c}{D_{tx} + D_{rx}}\frac{1}{p}, \label{eq:resolution}
\end{align}
where the over sampling factor $p$ can be used e.g. to reduce folding of frequencies $>f_c$. A $p$ factor of one has been used throughout this paper.

Equation (\ref{eq:resolution}) was used to calculate the object position or beam spacing in Fig.\,\ref{fig:das_capon_beams}. Therefore, if (\ref{eq:resolution}) is not fulfilled the scalloping loss could become visible in the image and the system will no longer be locally laterally shift-invariant. In an ultrasound image with 256 gray levels and 50 dB dynamic range, a -1 dB loss corresponds to 5 gray levels. This is approximately equal to the visibility threshold (Weber fraction of 2\%) for a grayscale image, and a loss larger than 1 dB could therefore end up being visible to the observer. In our further analysis a system would, based on this reasoning, be regarded as locally laterally shift-invariant if the loss in between beams is less than or close to -1 dB. %However, depending on viewing light conditions and e.g. the selected gray level mapping, a larger loss might be acceptable.

In order to visualize the local lateral shift-invariance of a given imaging system, we will make use of lateral shift-variance plots (LSV-plots), introduced by Hergum \textit{et al.} \cite{Hergum2007}. A LSV-plot is constructed for a given imaging system by imaging a point scatterer moving laterally at constant speed. For each lateral position the root-mean-square (RMS) of the image data is calculated in range, and the resulting beam profiles are stacked in to an image. For a given scan sequence there are twice as many point positions than there are receive lines. The LSV-plots in this paper are displayed as a contour plot with contours on -1, -2, -3, -6, -12 and -24 dB. The maximum for each beam profile is also marked with a black dot, and the amplitude variation among these points is presented in a subplot next to the LSV-plot. The LSV-plots based on simulations has been up-interpolated to 128 samples using cubic interpolation before presentation.

To further quantify the shift-invariance we have calculated the mean absolute error (MAE) of the maximum point position in each beam profile versus a straight line ($f(x)=x$). We will refer to this measure as P-MAE. The MAE of the maximum point's amplitude versus the data maximum is also measured. We refer to this measure as A-MAE. The first measure quantifies the amount of geometric distortion \cite{Hergum2007}, and the latter quantifies the peak gain variation or in other words the scalloping loss. The maximum amplitude deviation, or maximum scalloping loss, along this point trace is also calculated. We present all these numbers together with the LSV-plot.

A given imaging system will then be locally laterally shift-invariant if the contours are diagonal and if the amplitude deviation among the maximum points is less than or close to -1 dB. Then there is minimal geometric distortion (low P-MAE) and the scalloping loss will hopefully not be visible for the observer (low A-MAE). %Since both effects can occur at the same time it is important to have a measure not effected by geometric distortion.

Figure \ref{fig:das_shiftinvariant} presents such a LSV-plot for an imaging system using the DAS beamformer with a beam spacing equal to (\ref{eq:resolution}). The dotted vertical lines indicates the position of transmit-receive beam pairs. The point scatterer response has been simulated using Field II \cite{Jensen1992, Jensen1996a} with a probe setup equal to the M5S probe by GE Vingmed Ultrasound, Horten, Norway. The probe was excited with a 2.5 MHz and 1.5 cycles long pulse with a $\text{f-number}=3$ and corresponding cosine apodization on receive. The transmit focus in azimuth and elevation was configured to be equal and the point scatterer was moved through this focus. The lateral sampling density has been doubled prior to detection through coherent interpolation. We see that the contours in Fig.\,\ref{fig:das_shiftinvariant} are approximately diagonal (low P-MAE) and that the amplitude variation among the maximum points is minimal (low A-MAE). We therefore conclude that the system is locally laterally shift-invariant. In this, and the following LSV-plots based on simulations, the scan sequence always has a beam at zero degrees and the point scatterer is located exactly on this beam at one time instance.

The scalloping loss in Fig.\,\ref{fig:das_shiftinvariant} is smaller than what was predicted by Fig.\,\ref{fig:das_capon_beams}. To understand why note that Fig.\,\ref{fig:das_capon_beams} shows the scanned response for a narrowband system while the imaging system in Fig.\,\ref{fig:das_shiftinvariant} uses a broadband pulse. The beam in Fig.\,\ref{fig:das_shiftinvariant} will therefore be more similar to a squared gaussian function than the squared $sinc$ functions depicted in Fig.\,\ref{fig:das_capon_beams}. The imaging system in Fig.\,\ref{fig:das_shiftinvariant} also uses an expanding cosine window on receive.

@@

\begin{figure*}[!t]
\centerline{
\subfloat[]{\includegraphics[width=0.5\linewidth]{\img fig2a.eps}\label{fig:das_shiftinvariant}}
\hfill{}
\subfloat[]{\includegraphics[width=0.5\linewidth]{\img fig2b.eps}\label{fig:das_shiftvariant}}
}
\caption{LSV-plots of delay-and-sum beamforming. If the contours are diagonal and the amplitude variation among the peak points (black dots) is small, the imaging system is said to be locally laterally shift-invariant. Dotted vertical lines are plotted where transmit-receive beam pairs are located. a) Sampling density according to (\ref{eq:resolution}). The system is locally laterally shift-invariant. b) Undersampling with a factor of two. The system is no longer locally laterally shift-invariant.}
\label{fig:das}
\end{figure*}

\section{Local Lateral Shift-Invariance of the Capon Beamformer}\label{sec:capon_LLSI}
We will now investigate the local lateral shift-invariance of the Capon beamformer with parameters equal to the ones used in earlier work \cite{Synnevag2009, Asen}. In Fig.\,\ref{fig:capon1} the same point scatterer response as presented in Fig.\,\ref{fig:das_shiftinvariant} has been weighted with Capon weights calculated using (\ref{eq:capon_weights}) with $L = M/2 = 48$, $K=1$ and $d=1/100$. As mentioned, the set of lateral displacements is constructed in such a way that the center beam is the only one hitting the point scatterer exactly. The signal cancellation caused by the difference between the signal propagation vector and the assumed steering vector is therefore minimal at this location. From the plot we observe that the signal cancellation is high in all the other lateral positions. As a consequence the scalloping loss is tremendous, and not far from the predicted value in Fig\,\ref{fig:das_capon_beams}. %The difference between the predicted and the simulated loss might be caused by several factors. The most important one being that Fig.\,\ref{fig:das_capon_beams} shows a single frequency whereas the RMS operator used in Fig.\,\ref{fig:capon1} average a broadband signal in range. For a slice through the center of the pulse the scalloping loss is below -30dB.

@@

\begin{figure*}[!t]
\centerline{
\subfloat[]{
\includegraphics[width=0.5\linewidth]{\img fig3a.eps}\label{fig:capon1}
}
\subfloat[]{
\includegraphics[width=0.5\linewidth]{\img fig3b.eps}\label{fig:capon2}
}
}
\caption{LSV-plots of Capon beamforming with $L=M/2=48$ and $K=1$, and with sampling corresponding to (\ref{eq:resolution}). a) Capon beamforming with $d=1/100$ as diagonal loading. The system is not locally lateral shift-invariant. b) The diagonal loading factor $d$ is increased to 1. The system is still not locally lateral shift-invariant.}
\label{fig:capon}
\end{figure*}

@@

\subsection{Oversampling on Transmit}
The straight forward way of obtaining sufficient sampling in an active system is by increasing the number of transmit beams (tx-beams) until the system is locally laterally shift-invariant. In Fig.\,\ref{fig:capon_oversampling_a} and Fig.\,\ref{fig:capon_oversampling_b} we have increased the lateral sampling on transmit by a factor of four and ten respectively. As expected the geometric distortion and scalloping loss decreases with increased lateral sampling. Note that Fig.\,\ref{fig:capon_oversampling_b} is zoomed compared to Fig.\,\ref{fig:capon} in order to better see the beam-to-beam variation.

@@

\begin{figure*}[!t]
\centerline{
\subfloat[]{
\includegraphics[width=0.5\linewidth]{\img fig4a.eps}\label{fig:capon_oversampling_a}
}
\subfloat[]{
\includegraphics[width=0.5\linewidth]{\img fig4b.eps}\label{fig:capon_oversampling_b}
}
}
\caption{Capon beamforming as described in Fig\,\ref{fig:capon} with oversampling on transmit. The figure is zoomed compared with Fig\,\ref{fig:capon} to better see the beam-to-beam variation. a) Four times oversampling. b) Ten times oversampling. The maximum scalloping loss is still too high for the system to be called locally laterally shift invariant.}
\label{fig:capon_oversampling}
\end{figure*}

@@

For an uniform linear array located along the x-axis the steering vector $\vec{a}_\theta$ should be calculated as 
\begin{align}
\vec{a}_\theta = 
\begin{bmatrix}
e^{-j\frac{2\pi}{\lambda_c}x_0\sin(\theta)} \\
e^{-j\frac{2\pi}{\lambda_c}x_1\sin(\theta)} \\
\vdots \\ 
e^{-j\frac{2\pi}{\lambda_c}x_{L-1}\sin(\theta)}
\end{bmatrix}
\end{align}
where $x_i$ is the element position and $\theta$ is swiped from $-\Delta/2$ to $\Delta/2$ where $\Delta$ is as defined in (\ref{eq:resolution}). Note that the steering vector is calculated for a $L$ element array centered around the origin, even though the full array has $M$ elements. It is important that this subarray has the same element pitch, $x_i - x_{i-1}$, as the full array. The steering error introduced by this steering vector for a frequency band $B$ is less than $B/2 \times \Delta/2$.

@@

Since the phase rotation method operates on the same data as before, there will be no reduction in the acquisition frame rate by applying this method. In addition, since the same number of covariance matrices are constructed and inverted, there is no extra cost involved here as well. The only additional cost will be the multiplication of the extra steering vectors with the inverted covariance matrix. If the equation $\mat{\hat{R}}\vec{a} = \vec{b}$, as in \cite{Asen}, is solved using the Gauss-Jordan algorithm, we can solve the problem for all steering vectors at once by row reducing the matrix $[\mat{\hat{R}},  \vec{a}_{\theta_1}, \vec{a}_{\theta_2}, \cdots \vec{a}_{\theta_{N_\theta}}]$ to the following form $[\mat{I},  \vec{b}_{\theta_1}, \vec{b}_{\theta_2}, \cdots \vec{b}_{\theta_{N_\theta}}]$. This has lower complexity than inverting $\mat{\hat{R}}$, $O(L^3)$, as long as $N_\theta < L$.

@@

\section{In-vitro phantom data}\label{sec:res}
In Fig.\,\ref{fig:LSV_invitro} we present the result of applying the Capon beamformer on \textit{in-vitro} data with and without oversampling by phase rotation. The data was acquired using a modified Vivid E9 high-end cardiac ultrasound scanner by GE Vingmed Ultrasound, Horten, Norway, equipped with a M5S transducer operating in harmonic mode with a receive frequency of 3.7 MHz. The transmit beam density was set to match (\ref{eq:resolution}) based on this frequency. Note that in harmonic imaging the generated transmit beam at the harmonic frequency is a weighted version of the beam resulting from transmitting at the same frequency \cite{Fedewa2004}. The lateral sampling is therefore better than what was used in simulations. A tissue mimicking phantom containing wire targets and anechoic cysts was imaged, and the data was acquired while the probe was moved across the phantom manually. Fig.\,\ref{fig:image_das} shows the first frame out of 14 in the acquired loop. %Movies of the whole loop processed with DAS and Capon beamforming with and without ten times oversampling by phase rotation can be found in the supplementary material \multimedia{Videos}.
